---
title: "DATA 606 Lab 9"
subtitle: "Multiple linear regression"
author: "Donny Lofland"
output:
  html_document:
    css: ./lab.css
    highlight: pygments
    theme: cerulean
  pdf_document: default
---

## Grading the professor

Many college courses conclude by giving students the opportunity to evaluate 
the course and the instructor anonymously. However, the use of these student 
evaluations as an indicator of course quality and teaching effectiveness is 
often criticized because these measures may reflect the influence of 
non-teaching related characteristics, such as the physical appearance of the 
instructor. The article titled, "Beauty in the classroom: instructors' 
pulchritude and putative pedagogical productivity" (Hamermesh and Parker, 2005) 
found that instructors who are viewed to be better looking receive higher 
instructional ratings. (Daniel S. Hamermesh, Amy Parker, Beauty in the 
classroom: instructors pulchritude and  putative pedagogical productivity, 
*Economics of Education Review*, Volume 24, Issue 4, August 2005, Pages 369-376, 
ISSN 0272-7757, 10.1016/j.econedurev.2004.07.013. [http://www.sciencedirect.com/science/article/pii/S0272775704001165](http://www.sciencedirect.com/science/article/pii/S0272775704001165).)

In this lab we will analyze the data from this study in order to learn what goes 
into a positive professor evaluation.

## The data

The data were gathered from end of semester student evaluations for a large 
sample of professors from the University of Texas at Austin. In addition, six 
students rated the professors' physical appearance. (This is aslightly modified 
version of the original data set that was released as part of the replication 
data for *Data Analysis Using Regression and Multilevel/Hierarchical Models* 
(Gelman and Hill, 2007).) The result is a data frame where each row contains a 
different course and columns represent variables about the courses and professors.

```{r load-data, eval=TRUE}
load("more/evals.RData")
```

variable         | description
---------------- | -----------
`score`          | average professor evaluation score: (1) very unsatisfactory - (5) excellent.
`rank`           | rank of professor: teaching, tenure track, tenured.
`ethnicity`      | ethnicity of professor: not minority, minority.
`gender`         | gender of professor: female, male.
`language`       | language of school where professor received education: english or non-english.
`age`            | age of professor.
`cls_perc_eval`  | percent of students in class who completed evaluation.
`cls_did_eval`   | number of students in class who completed evaluation.
`cls_students`   | total number of students in class.
`cls_level`      | class level: lower, upper.
`cls_profs`      | number of professors teaching sections in course in sample: single, multiple.
`cls_credits`    | number of credits of class: one credit (lab, PE, etc.), multi credit.
`bty_f1lower`    | beauty rating of professor from lower level female: (1) lowest - (10) highest.
`bty_f1upper`    | beauty rating of professor from upper level female: (1) lowest - (10) highest.
`bty_f2upper`    | beauty rating of professor from second upper level female: (1) lowest - (10) highest.
`bty_m1lower`    | beauty rating of professor from lower level male: (1) lowest - (10) highest.
`bty_m1upper`    | beauty rating of professor from upper level male: (1) lowest - (10) highest.
`bty_m2upper`    | beauty rating of professor from second upper level male: (1) lowest - (10) highest.
`bty_avg`        | average beauty rating of professor.
`pic_outfit`     | outfit of professor in picture: not formal, formal.
`pic_color`      | color of professor's picture: color, black & white.

## Exploring the data

1.  Is this an observational study or an experiment? The original research
    question posed in the paper is whether beauty leads directly to the
    differences in course evaluations. Given the study design, is it possible to
    answer this question as it is phrased? If not, rephrase the question.

> This is an observational study.  We cannot easily answer the question as phrased ... to isolate and test whether beauty (farily subjective) causes changes in course evalutaion would require randomized trials.  In this case, we can more appropriately ask and answer is there a correlation between beauty and evaluations and/o how much of the variablility in course evaluations might be explained by beauty.

2.  Describe the distribution of `score`. Is the distribution skewed? What does 
    that tell you about how students rate courses? Is this what you expected to 
    see? Why, or why not?

```{r}
hist(evals$score)
```

> Scores are left skewed where the most scores are between 4.0 and 5.0 and a long tail going down to 2.0.  In general students give courses above average scores >= 3.5.  An ideal scoring prototcal would have resulted in a more normalized curve with the bulk in the middle and symmetic tails in both directions.  This is not surprising ... I would have expected more courses to receive good score than bad score - if a cource or professor scored poorly in previous semesters, the would have likely been changes to improve scores.  So socre don't really represent a random sample ... we would expect over time to weed out pooring performing classes/materials and/or professors. 

3.  Excluding `score`, select two other variables and describe their relationship 
    using an appropriate visualization (scatterplot, side-by-side boxplots, or 
    mosaic plot).

```{r}
plot(evals$rank, evals$age)
```

> Looking at rank vs age, we would expect tenure track faculty to be younger than tenured faculty and expect tenured.

## Simple linear regression

The fundamental phenomenon suggested by the study is that better looking teachers
are evaluated more favorably. Let's create a scatterplot to see if this appears 
to be the case:

```{r scatter-score-bty_avg, eval = TRUE}
plot(evals$score ~ evals$bty_avg)
```


Before we draw conclusions about the trend, compare the number of observations 
in the data frame with the approximate number of points on the scatterplot. 
Is anything awry?

```{r}
nrow(evals)
```

> There are 463 rows in the data set, but far fewer points.

4.  Replot the scatterplot, but this time use the function `jitter()` on the
    $y$- or the $x$-coordinate. (Use `?jitter` to learn more.) What was
    misleading about the initial scatterplot?

```{r scatter-score-bty_avg_jit, eval = TRUE}
plot(jitter(evals$score) ~ evals$bty_avg)
```

> The initial scatter plot had overlapping points.  With jitter we can more clearly see where the bulk of points are landing.

5.  Let's see if the apparent trend in the plot is something more than
    natural variation. Fit a linear model called `m_bty` to predict average
    professor score by average beauty rating and add the line to your plot
    using `abline(m_bty)`. Write out the equation for the linear model and
    interpret the slope. Is average beauty score a statistically significant
    predictor? Does it appear to be a practically significant predictor?

```{r}
m_bty <- lm(evals$score ~ evals$bty_avg)
summary(m_bty)

plot(jitter(evals$score) ~ evals$bty_avg)
abline(m_bty)
```

> $score = 3.88034 + 0.06664 * bty_avg$

> As avg_bty increases, the scores also increaeses; while the slope and intercept are "significant" (ie there is a positive correlation), the R$^2$ is ~0.033 meaning only about 3.5% of the variation in score can be explained by beauty.  ~96.7% of the variation is due to other factors and/or randomness.

6.  Use residual plots to evaluate whether the conditions of least squares
    regression are reasonable. Provide plots and comments for each one (see
    the Simple Regression Lab for a reminder of how to make these).
    
```{r}

plot(m_bty$residuals ~ evals$bty_avg, 
     xlab = "Beauty", ylab = "Residuals", 
     pch = 19, 
     axes = FALSE)
axis(1, at = seq(-1, 2, 1))
axis(2, at = seq(-1, 1, 1))
box()
abline(h = 0, lty = 3)

hist(m_bty$residuals, 
     xlab = "Residuals", ylab = "", main = "",
     xlim = c(-2,2))

qqnorm(m_bty$residuals, 
       pch = 19, 
       main = "", las = 0)
qqline(m_bty$residuals)

plot(m_bty$residuals, 
     xlab = "Order of data collection", ylab = "Residuals", main = "",
     pch = 19, 
     ylim = c(-1.82, 1.82), axes = FALSE)
axis(1)
axis(2, at = seq(-1, 1, 1))
box()
abline(h = 0, lty = 3)
```
    
> Residuals while not perfectly normally distributed (left skewed a little), do appear to be overall mostly normal.  There so not appear to be any trends.  The assumptions are broadly met. 

## Multiple linear regression

The data set contains several variables on the beauty score of the professor: 
individual ratings from each of the six students who were asked to score the 
physical appearance of the professors and the average of these six scores. Let's 
take a look at the relationship between one of these scores and the average 
beauty score.

```{r bty-rel, eval = TRUE}
plot(evals$bty_avg ~ evals$bty_f1lower)
cor(evals$bty_avg, evals$bty_f1lower)
```

As expected the relationship is quite strong - after all, the average score is 
calculated using the individual scores. We can actually take a look at the 
relationships between all beauty variables (columns 13 through 19) using the 
following command:

```{r bty-rels, eval = TRUE}
plot(evals[,13:19])
```

These variables are collinear (correlated), and adding more than one of these 
variables to the model would not add much value to the model. In this 
application and with these highly-correlated predictors, it is reasonable to use
the average beauty score as the single representative of these variables.

In order to see if beauty is still a significant predictor of professor score 
after we've accounted for the gender of the professor, we can add the gender 
term into the model.

```{r scatter-score-bty_avg_gender, eval = TRUE}
m_bty_gen <- lm(score ~ bty_avg + gender, data = evals)
summary(m_bty_gen)
```

7.  P-values and parameter estimates should only be trusted if the
    conditions for the regression are reasonable. Verify that the conditions
    for this model are reasonable using diagnostic plots.

```{r}
plot(m_bty_gen)
```

```{r}
par(mfrow = c(2, 2))

plot(m_bty_gen$residuals ~ evals$bty_avg, 
     xlab = "Beauty", ylab = "Residuals", 
     pch = 19,  
     axes = FALSE)
axis(1, at = seq(-1, 2, 1))
axis(2, at = seq(-1, 1, 1))
box()
abline(h = 0, lty = 3)

hist(m_bty_gen$residuals, 
     xlab = "Residuals", ylab = "", main = "",
     xlim = c(-2,2))

qqnorm(m_bty_gen$residuals, 
       pch = 19,
       main = "", las = 0)
qqline(m_bty_gen$residuals)

plot(m_bty_gen$residuals, 
     xlab = "Order of data collection", ylab = "Residuals", main = "",
     pch = 19, 
     ylim = c(-1.82, 1.82), axes = FALSE)
axis(1)
axis(2, at = seq(-1, 1, 1))
box()
abline(h = 0, lty = 3)

```

> Assumptions seem reasonable: linearity of data, residuals are normal, no patterns in residuls, no strong leverage points.

8.  Is `bty_avg` still a significant predictor of `score`? Has the addition
    of `gender` to the model changed the parameter estimate for `bty_avg`?

> Bty_avg is still significant and together with gender we now explain 5.5% of the variation in scores.  The presence of gender has improved our model slightly, but while these are significant features, they offer low explanatory value.

Note that the estimate for `gender` is now called `gendermale`. You'll see this 
name change whenever you introduce a categorical variable. The reason is that R 
recodes `gender` from having the values of `female` and `male` to being an 
indicator variable called `gendermale` that takes a value of $0$ for females and
a value of $1$ for males. (Such variables are often referred to as "dummy" 
variables.)

As a result, for females, the parameter estimate is multiplied by zero, leaving 
the intercept and slope form familiar from simple regression.

\[
  \begin{aligned}
\widehat{score} &= \hat{\beta}_0 + \hat{\beta}_1 \times bty\_avg + \hat{\beta}_2 \times (0) \\
&= \hat{\beta}_0 + \hat{\beta}_1 \times bty\_avg\end{aligned}
\]

We can plot this line and the line corresponding to males with the following 
custom function.

```{r twoLines, eval = TRUE}
multiLines(m_bty_gen)
```

9.  What is the equation of the line corresponding to males? (*Hint:* For
    males, the parameter estimate is multiplied by 1.) For two professors
    who received the same beauty rating, which gender tends to have the
    higher course evaluation score?

> $score = (3.74734 + 0.17239) + 0.07416 * bty_avg$

The decision to call the indicator variable `gendermale` instead of`genderfemale`
has no deeper meaning. R simply codes the category that comes first 
alphabetically as a $0$. (You can change the reference level of a categorical 
variable, which is the level that is coded as a 0, using the`relevel` function. 
Use `?relevel` to learn more.)

10. Create a new model called `m_bty_rank` with `gender` removed and `rank` 
    added in. How does R appear to handle categorical variables that have more 
    than two levels? Note that the rank variable has three levels: `teaching`, 
    `tenure track`, `tenured`.

```{r}
m_bty_rank <- lm(score ~ bty_avg + rank, data = evals)
summary(m_bty_rank)
```

> R creates a separate value for each rank, again leaving off the first alphabetic category which is treated as 0.  Depending on which rank we are interested in, we use that value and the other is multiplied by zero so it drops out.

The interpretation of the coefficients in multiple regression is slightly 
different from that of simple regression. The estimate for `bty_avg` reflects
how much higher a group of professors is expected to score if they have a beauty
rating that is one point higher *while holding all other variables constant*. In
this case, that translates into considering only professors of the same rank 
with `bty_avg` scores that are one point apart.

## The search for the best model

We will start with a full model that predicts professor score based on rank, 
ethnicity, gender, language of the university where they got their degree, age, 
proportion of students that filled out evaluations, class size, course level, 
number of professors, number of credits, average beauty rating, outfit, and 
picture color.

11. Which variable would you expect to have the highest p-value in this model? 
    Why? *Hint:* Think about which variable would you expect to not have any 
    association with the professor score.

> I might expect the following to be significant
* *language* (communication ability) 
* *age* (experience level)
* *cls_credits* (time spent in class with professor)

Let's run the model...

```{r m_full, eval = TRUE, tidy = FALSE}
m_full <- lm(score ~ rank + ethnicity + gender + language + age + cls_perc_eval 
             + cls_students + cls_level + cls_profs + cls_credits + bty_avg 
             + pic_outfit + pic_color, data = evals)
summary(m_full)
```

12. Check your suspicions from the previous exercise. Include the model output
    in your response.

> Language and age were significant as expected.  I was surprised that the percent of student completing an evaluation was significant along with whether the picture was color or black and white.

13. Interpret the coefficient associated with the ethnicity variable.

> Non-minority professors scored 0.21 higher than minority professors.  

14. Drop the variable with the highest p-value and re-fit the model. Did the
    coefficients and significance of the other explanatory variables change?
    (One of the things that makes multiple regression interesting is that
    coefficient estimates depend on the other variables that are included in
    the model.) If not, what does this say about whether or not the dropped
    variable was collinear with the other explanatory variables?

```{r}
m_full <- lm(score ~ rank + ethnicity + gender + language + age + cls_perc_eval 
             + cls_students + cls_level + cls_credits + bty_avg 
             + pic_outfit + pic_color, data = evals)
summary(m_full)
```

> I dropped cls_profs (0.77806).  Overall, the coefficients and significance did NOT change by much.  This suggests the dropped variable did not add any unqiue explanitory value that wasn't alrady accounted for by other features.

15. Using backward-selection and p-value as the selection criterion,
    determine the best model. You do not need to show all steps in your
    answer, just the output for the final model. Also, write out the linear
    model for predicting score based on the final model you settle on.

```{r}
m_full <- lm(score ~ ethnicity + gender + language + age + cls_perc_eval 
             + cls_credits + bty_avg 
             + pic_color, data = evals)
summary(m_full)
```

> $score = 3.771922 + (not_minority * 0.167872) + (male * 0.207112) - (non_english * 0.206178) - (age * 0.006046) + (cls_perc_eval * 0.004656) + (one_credit * 0.505306) + (byt_avg * 0.051069) - (color_pic * 0.190579)  $

> where not_minority, male, non_english, one_credit, and color_pic are boolean 0 or 1 depending on whether its that factor or not.

16. Verify that the conditions for this model are reasonable using diagnostic 
    plots.

```{r}
par(mfrow = c(2, 2))
plot(m_full)
```

```{r}

hist(m_full$residuals, 
     xlab = "Residuals", ylab = "", main = "",
     xlim = c(-2,2))

qqnorm(m_full$residuals, 
       pch = 19, 
       main = "", las = 0)
qqline(m_bty_gen$residuals)

plot(m_full$residuals, 
     xlab = "Order of data collection", ylab = "Residuals", main = "",
     pch = 19,
     ylim = c(-1.82, 1.82), axes = FALSE)
axis(1)
axis(2, at = seq(-1, 1, 1))
box()
abline(h = 0, lty = 3)
```

> The conditions for this model seem reasonable: residuals look fine, a linear model fits and there are no really problematic leverage points.

17. The original paper describes how these data were gathered by taking a
    sample of professors from the University of Texas at Austin and including 
    all courses that they have taught. Considering that each row represents a 
    course, could this new information have an impact on any of the conditions 
    of linear regression?

> We may be skewing our results if a single professor teaches many courses comapred with those that only tought one.  We were assuming independence between observation, but if a single professor shows up multple times, those rows are no longer independent.

18. Based on your final model, describe the characteristics of a professor and 
    course at University of Texas at Austin that would be associated with a high
    evaluation score.

> White, male, english speaking, younger, good looking, teaching a one-credit course and whose photo was in black & white.

19. Would you be comfortable generalizing your conclusions to apply to professors
    generally (at any university)? Why or why not?

> Nope.  Students are doing the evaluations and we cannot assume students at one univeristy would evaluate professors the same as at different university.  We also didn't explore causality, merely correlations.  Evaluations of professors may also be affected by what other faculty the students are exposed to, what courses students take, the types of classes, difficulty of classes, etc.  As an observational study we are limited to stating very simple correlations and that they only hold for the evaluated courses & professors.  In fact, at the same university, these patterns might not hold in other departments.  Our best model only accounted for a very small amount of the variability, so there could be many other factors we didnt' measure that could play more important roles.
