---
title: 'DATA606 Lab 7'
subtitle: 'Inference for numerical data'
author: 'Donny Lofland'
data: '10/26/2019'
output:
  html_document:
    theme: cerulean
    highlight: pygments
    css: ./lab.css
    toc: true
    toc_float: true
  pdf_document: default
---

Source files: [https://github.com/djlofland/DATA606_F2019/tree/master/Lab7](https://github.com/djlofland/DATA606_F2019/tree/master/Lab7)

```{r setup, eval=TRUE}
library(tidyverse)
```

## North Carolina births

In 2004, the state of North Carolina released a large data set containing 
information on births recorded in this state. This data set is useful to 
researchers studying the relation between habits and practices of expectant 
mothers and the birth of their children. We will work with a random sample of 
observations from this data set.

## Exploratory analysis

Load the `nc` data set into our workspace.

```{r load-data, eval=TRUE}
load("more/nc.RData")
```

We have observations on 13 different variables, some categorical and some 
numerical. The meaning of each variable is as follows.

variable         | description
---------------- | -----------
`fage`           | father's age in years.
`mage`           | mother's age in years.
`mature`         | maturity status of mother.
`weeks`          | length of pregnancy in weeks.
`premie`         | whether the birth was classified as premature (premie) or full-term.
`visits`         | number of hospital visits during pregnancy.
`marital`        | whether mother is `married` or `not married` at birth.
`gained`         | weight gained by mother during pregnancy in pounds.
`weight`         | weight of the baby at birth in pounds.
`lowbirthweight` | whether baby was classified as low birthweight (`low`) or not (`not low`).
`gender`         | gender of the baby, `female` or `male`.
`habit`          | status of the mother as a `nonsmoker` or a `smoker`.
`whitemom`       | whether mom is `white` or `not white`.

### Exercise 1

1.  What are the cases in this data set? How many cases are there in our sample?

> Each case in the dataset (row) is **a single birth event**.  There are **`r nrow(nc)` cases** in the dataset.  

As a first step in the analysis, we should consider summaries of the data. This
can be done using the `summary` command:

```{r summary, eval=TRUE}
summary(nc)
```

As you review the variable summaries, consider which variables are categorical 
and which are numerical. For numerical variables, are there outliers? If you 
aren't sure or want to take a closer look at the data, make a graph.

> Categorical: **mature, premie, marital, lowbirthweight, gender, habit, whitemom**

> Numerical: **fage, mage, weeks, visits, gained, weight **

```{r outliers}
boxplot(nc$fage, nc$mage, nc$weeks, nc$visits, nc$gained, nc$weight)
```

> fage as 2 high outliers, mage has 1 high outlier, weeks has a number of low outliers and 1 high outlier and has a clear right-skew, visits has outliers both above and below the interquartile range with a lisght left skew towards fewer visits, gained has a number of high outliers but those aside, looks fairly "normal", and weight has a number of outliers.

Consider the possible relationship between a mother's smoking habit and the 
weight of her baby. Plotting the data is a useful first step because it helps 
us quickly visualize trends, identify strong associations, and develop research
questions.

### Exercise 2

2.  Make a side-by-side boxplot of `habit` and `weight`. What does the plot 
highlight about the relationship between these two variables?

```{r habit_weight, eval=TRUE}
boxplot(nc$weight~nc$habit)
```

> Nonsmoker baby weight has a much wider range of values wit more outliers above and below the IQR.  Non-smokers  baby weights appear to be slightly higher on average.  Smokers' baby weights appear to be slightly lower with a narrower range of values.

The box plots show how the medians of the two distributions compare, but we can
also compare the means of the distributions using the following function to 
split the `weight` variable into the `habit` groups, then take the mean of each
using the `mean` function.

```{r by-means, eval=TRUE}
by(nc$weight, nc$habit, mean)
```

There is an observed difference, but is this difference statistically 
significant? In order to answer this question we will conduct a hypothesis test.

## Inference

### Exercise 3

3.  Check if the conditions necessary for inference are satisfied. Note that 
you will need to obtain sample sizes to check the conditions. You can compute 
the group size using the same `by` command above but replacing `mean` with 
`length`.

```{r exercise3, eval=TRUE}
by(nc$weight, nc$habit, length)
```

> Conditions: cases were randomly sampled. We have > 10 cases in each group, smoker and nonsmoker 

### Exercise 4

4.  Write the hypotheses for testing if the average weights of babies born to 
smoking and non-smoking mothers are different.

> $H_{0}$: There is **no difference** between baby weights based on mom smoking habit

> $H_{A}$: There **is** as differnce between the baby weights of moms who smoke vs those who don't

Next, we introduce a new function, `inference`, that we will use for conducting
hypothesis tests and constructing confidence intervals. 

```{r inf-weight-habit-ht, eval=TRUE, tidy=FALSE}
inference(y = nc$weight, x = nc$habit, est = "mean", type = "ht", null = 0, 
          alternative = "twosided", method = "theoretical")
```

Let's pause for a moment to go through the arguments of this custom function. 
The first argument is `y`, which is the response variable that we are 
interested in: `nc$weight`. The second argument is the explanatory variable, 
`x`, which is the variable that splits the data into two groups, smokers and 
non-smokers: `nc$habit`. The third argument, `est`, is the parameter we're 
interested in: `"mean"` (other options are `"median"`, or `"proportion"`.) Next
we decide on the `type` of inference we want: a hypothesis test (`"ht"`) or a 
confidence interval (`"ci"`). When performing a hypothesis test, we also need 
to supply the `null` value, which in this case is `0`, since the null 
hypothesis sets the two population means equal to each other. The `alternative`
hypothesis can be `"less"`, `"greater"`, or `"twosided"`. Lastly, the `method` 
of inference can be `"theoretical"` or `"simulation"` based.

### Exercise 5

5.  Change the `type` argument to `"ci"` to construct and record a confidence 
interval for the difference between the weights of babies born to smoking and 
non-smoking mothers.

By default the function reports an interval for ($\mu_{nonsmoker} - \mu_{smoker}$)
. We can easily change this order by using the `order` argument:

```{r inf-weight-habit-ci, eval=TRUE, tidy=FALSE}
inference(y = nc$weight, x = nc$habit, est = "mean", type = "ci", null = 0, 
          alternative = "twosided", method = "theoretical", 
          order = c("smoker","nonsmoker"))
```

* * *

## On your own

### Problem 1

1. Calculate a 95% confidence interval for the average length of pregnancies 
(`weeks`) and interpret it in context. Note that since you're doing inference 
on a single population parameter, there is no explanatory variable, so you can 
omit the `x` variable from the function.

```{r oyo1}
inference(y = nc$weeks, est = "mean", type = "ci", null = 0, 
          alternative = "twosided", method = "theoretical")
```

> **( 38.1528 , 38.5165 )**

### Problem 2

2. Calculate a new confidence interval for the same parameter at the 90% 
confidence level. You can change the confidence level by adding a new argument 
to the function: `conflevel = 0.90`.

```{r oyo2}
inference(y = nc$weeks, est = "mean", type = "ci", null = 0, 
          alternative = "twosided", method = "theoretical",
          conflevel = 0.90)
```

> **( 38.182 , 38.4873 )**

### Problem 3

3. Conduct a hypothesis test evaluating whether the average weight gained by 
younger mothers is different than the average weight gained by mature mothers.

> Conditions: 867 younger women and 133 mature women

> $H_{0}$: Weight gained between young and mature women is the same

> $H_{A}$: Weight gained between young and mature women is different.

```{r oyo3}
inference(y = nc$gained, x=nc$mature, est = "mean", type = "ht", null = 0, 
          alternative = "twosided", method = "theoretical",
          )

inference(y = nc$gained, x=nc$mature, est = "mean", type = "ci", null = 0, 
          alternative = "twosided", method = "theoretical",
          )
```

> While the mean weight is lower for mature women, it is not statistically significant at the 95% confidence.  0 lbs (our null hypothesis) falls within the 95% confidence interval (-4.2896 ,0.7502) so we cannot reject the null hypothesis.

### Problem 4

4.  Now, a non-inference task: Determine the age cutoff for younger and mature 
mothers. Use a method of your choice, and explain how your method works.

```{r oyo4}
nc %>% group_by(mature) %>% summarize(min_age = min(mage), max_age=max(mage))
```

> youger women: **13-34** and mature women: **35-50** with the **cuttoff at 35yro**

### Problem 5

5.  Pick a pair of numerical and categorical variables and come up with a 
research question evaluating the relationship between these variables. 
Formulate the question in a way that it can be answered using a hypothesis test
and/or a confidence interval. Answer your question using the `inference` 
function, report the statistical results, and also provide an explanation in 
plain language.

> **Question: Is there a difference in birth weights between male and female babies?**

```{r oyo5}
qqnorm(nc$weight[nc$gender=='male'])
qqline(nc$weight[nc$gender=='male'])

qqnorm(nc$weight[nc$gender=='female'])
qqline(nc$weight[nc$gender=='female'])

inference(y = nc$weight, x=nc$gender, est = "mean", type = "ht", null = 0, 
          alternative = "twosided", method = "theoretical",
          )

inference(y = nc$weight, x=nc$gender, est = "mean", type = "ci", null = 0, 
          alternative = "twosided", method = "theoretical",
          )
```

> The p-value, 0, from the hypotheis test suggest that we reject the null hypothesis (male and female babies have the same mean birth weight) and we should accept that the weights are different.  Since the weights are both skewed from normal, I'd lke to also explore whether the median weights are different between male and female babies.  

```{r oyo5-tdist}
inference(y = nc$weight, x=nc$gender, est = "median", type = "ht", null = 0, 
          alternative = "twosided", method = "simulation",
          )

inference(y = nc$weight, x=nc$gender, est = "median", type = "ci", null = 0, 
          alternative = "twosided", method = "simulation",
          )
```

> Looking at median weights, we also see a clear difference in the same direction with male babies weighting more at birth (p=0.0026).
